---
title: "5-TimeSeries2"
author: "Carson Batchelor"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Utilize the time series data of air-carrier freight for the analysis by following these steps:

```{r}
rm(list=ls())
air = ts(scan("../../data/AirFreight.txt"), start=c(1969,1), end=c(1978,12), frequency=12)
air
```


Step 1. Display the time series plot along with the corresponding ACF and PACF plots. a) Does the data exhibit a non-seasonal trend? b) Does the data exhibit a seasonality?

- Yes, the data exhibits an upward non-seasonal trend, as seen in the general increase in freight volume over time.

- Yes, the data exhibits seasonality with a period of 12 months, as seen in the repeating patterns within each year. in the acf() and pacf() plots, there are significant spikes at lags 0.0,1.0,2.0.

```{r}
# Time Series plot
ts.plot(air)

# Auto Correlation and Partial Correlation Plots
acf(air, lag.max=36)

pacf(air, lag.max=36)
```


Step 2. If your answer to 1-a) is No, skip this question. If your answer to 1-a) is Yes, propose a differencing method to remove the trend from the data. Provide the time series plot of the detrended data along with the corresponding ACF and PACF plots.

```{r}
# first difference
air_diff <- diff(air, lag=1)

# First Difference Time Series plot
ts.plot(air_diff)

# First Difference Auto Correlation and Partial Correlation Plots
acf(air_diff, lag.max=36)
pacf(air_diff, lag.max=36)
```

Step 3. If your answer to 1-b) is No, skip this question. If your answer to 1-b) is Yes, propose a differencing method to further eliminate seasonality. Provide the time series plot of the data with both trend (if applicable) and seasonality removed, along with the corresponding ACF and PACF plots.

```{r}
# Seasonal differencing to remove seasonality
air_diff_seas <- diff(air_diff, lag=12)

# Time series plot of data with trend and seasonality removed
ts.plot(air_diff_seas)

# ACF and PACF plots
acf(air_diff_seas, lag.max=36)
pacf(air_diff_seas, lag.max=36)
```

Step 4. Based on the analysis in steps 2 and 3, use auto.arima to assist model development. Fit and report your final model.

```{r}
library(forecast)

# Use auto.arima to select the best SARIMA model
final_model <- auto.arima(air_diff_seas, seasonal=TRUE)

# Report the final model
summary(final_model)
```

Step 5. Show the ACF and PACF plots of the residuals from your final model. Assess whether these plots suggest stationarity in the residuals.

- Yese these plots suggest stationarity because:
- The ACF and PACF plots show no significant autocorrelation with no spikes outside confidence bands.
- The residuals fluctuate around a constant mean (zero) with constant variance.

```{r}
# train best model on raw data
best_model <- Arima(air, order = c(0,1,1), seasonal = c(0,1,1), include.constant = TRUE)
# Extract residuals
residuals <- residuals(best_model)

# ACF and PACF plots of residuals
acf(residuals, lag.max=36)
pacf(residuals, lag.max=36)
```

Step 6. Utilize the final model to generate 36-month ahead forecasts for air-carrier freight. Plot both the forecasts and the forecast intervals.

```{r}
# Generate 36-month ahead forecasts
forecasts <- forecast(best_model, h=36)

# Plot the forecasts with forecast intervals
plot(forecasts, main="36-Month Forecasts for Air Freight", ylab="Freight Volume", xlab="Year")
```



