---
title: "GLMM Homework"
author: "Carson Batchelor"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Consider dataset PropertyFund. The goal is to predict whether a policyholder will have claims for a given year. Use data from years 2006 to 2009 for training and data from year 2010 for testing.

```{r, results = "hide", message = FALSE}
# Inputs
library(dplyr)
library(ggplot2)
library(tidyr)
library(lme4)
library(ROCR)
```

```{r}
# load property fund data
rm(list=ls())
pfDat = read.table("../../data/PropertyFund.txt", sep ="", quote = "",header=TRUE)
names(pfDat)

# create a binary outcome to indicate occurrence of claims
pfDat$Freq0 <- 1*(pfDat$Freq>0)

# split data into train and test
idx <- which(pfDat$Year < 2010)
traindat <- pfDat[idx,]
testdat <- pfDat[-idx,]
```

There are two analyst HOM and HET, and both use log(Coverage), log(Deduct), and EntityType as input variables in their model development. The analyst explores several modeling strategies:

1. Analyst HOM considers two homogenenous logistic models: (1) treating EntityType as fixed; (2) treating EntityType as random.

```{r}
# Homogeneous Logistic models
# Fixed effects model
logitFix <- glm(Freq0 ~ log(Coverage) + log(Deduct) + EntityType, family=binomial, data=traindat)

summary(logitFix)

# Random effects model
logitRE <- glmer(Freq0 ~ log(Coverage) + log(Deduct) + (1|EntityType), family=binomial, data=traindat)

summary(logitRE)
```

- 1a. Which model will analyst HOM select based on the goodness-of-fit statistics AIC.

  - Model 1 Fixed: AIC = 3964.7
  - Model 2 Random: AIC = 3981.0
  - Based on AIC, analyst HOM will choose model 1 that is the fixed effect model.
  
- 1b. which model will analyst HOM select based on the test error rate.

```{r}
# predictions
predFixed <- predict(logitFix, newdata = testdat, type="response")
predRE <- predict(logitRE, newdata = testdat, type="response")

# MSE
mseFixed <- mean((predFixed - testdat$Freq)^2) 
mseRE <- mean((predRE - testdat$Freq)^2) 
print(c(mseFixed,mseRE))

# Overall error rate
classFixed <- ifelse(predFixed > 0.5, 1, 0)
classRE <- ifelse(predRE > 0.5, 1, 0)

errFixed <- mean(classFixed != testdat$Freq0)
errRE <- mean(classRE != testdat$Freq0)

print(c(errFixed,errRE))
```
  - analyst HOM would select the fixed effect model 1 becuase of a lower MSE and lower test error rate

2. Analyst HET considers two heterogeneous logistic models: (1) treating EntityType as fixed; (2) treating EntityType as random.

```{r}
# Heterogeneous logistic fixed effect model
logitHetFix <- glm(Freq0 ~ log(Coverage) * EntityType + log(Deduct) * EntityType, 
                   family=binomial, data=traindat)
summary(logitHetFix)

# Heterogeneous Logistic Random effect model
logitHetRE <- glmer(Freq0 ~ log(Coverage) + log(Deduct) + (log(Coverage) + log(Deduct) | EntityType), family=binomial, data=traindat)

summary(logitHetRE)

```

- a.  Which model will analyst HET select based on the goodness-of-fit statistics AIC.

```{r}
AIC(logitHetFix)
AIC(logitHetRE)
```
  - Model 1 Fixed: AIC = 3963.18
  - Model 2 Random: AIC = 3979.515
  - Based on AIC, analyst HET will choose model 1 that is the fixed effect model.
  
- b.  which model will analyst HET select based on the test error rate.

```{r}
predHetFix <- predict(logitHetFix, newdata = testdat, type="response")
predHetRE <- predict(logitHetRE, newdata = testdat, type="response")

mseHetFix <- mean((predHetFix - testdat$Freq)^2)
mseHetRE <- mean((predHetRE - testdat$Freq)^2)

print(c(mseHetFix, mseHetRE))

# Overall error rate
classHetFixed <- ifelse(predHetFix > 0.5, 1, 0)
classHetRE <- ifelse(predHetRE > 0.5, 1, 0)

errHetFixed <- mean(classHetFixed != testdat$Freq0)
errHetRE <- mean(classHetRE != testdat$Freq0)

print(c(errHetFixed,errHetRE))
```
  - analyst HET would select the Random effect model 2 becuase of a lower test error rate
  
Compare the AUC on test data between the models selected by the two analysts based on test error rate.

  - Higher AUC better

```{r}
# Predictions for HOM’s chosen model (Fixed Effects Model)
predHOM <- predict(logitFix, newdata=testdat, type="response")

# Predictions for HET’s chosen model (Fixed Effects Heterogeneous Model)
predHET <- predict(logitHetRE, newdata=testdat, type="response")

# Convert actual labels to factor
actual <- testdat$Freq0  

# Function to compute AUC
compute_auc <- function(preds, actual) {
  pred_obj <- prediction(preds, actual)
  perf_obj <- performance(pred_obj, measure="auc")
  return(as.numeric(perf_obj@y.values))
}

# Compute AUC for both models
auc_HOM <- compute_auc(predHOM, actual)
auc_HET <- compute_auc(predHET, actual)

# Print AUC values
print(c(AUC_HOM = auc_HOM, AUC_HET = auc_HET))

```